working title Compatibility system and stygma size of pollen recipient as main predictors of heterospecific pollen effect

### 2018-12-04 Manuscript_draft/Manuscript_draft.Rmd
last edits

Recent studies have advanced in the ecological understanding of heterospecific pollen effect @morales2008 @ashman2013 @arceo2016. A general overview of foreign pollen arrival is that it can play an important role on species fitness but seems to be context dependent and not always produce a decrease in fitness @morales2008. Part of this unpredictability is due to the enormous variability of foreing pollen transferred in nature, where levels between 0 and 75 percent are seen, but most commonly values ranges between 0 and 20 percent of the total pollen load, being the generalist species the ones that receive greater loads of heterospecific pollen @bartomeus2008 @montgomery2012 @ashman2013 @fang2013. Although heterospecific pollen quantity is fundamental to understand the outcome of the interaction so is the different traits of both pollen donor and recipient. @ashman2013 postulated the first predictive framework for traits of heterospecific pollen effect, where different traits such as compatibility system and pollen size among others seems to be crucial to understand foreing pollen effect. Moreover, in @tong2016 an assymetric effect was shown in a crossing experiment between 6 species of the genus *Pedicularis* where the pollen of long styled species was able to grow the full length of the style on short styled species but not viceversa. Despite these recent caveats, we still lack empirical evidence to affirm what are the main traits that drive heterospecific pollen effect for both pollen donor and recipient at seed production level. Interestingly, to comprehend how these traits interact is also crucial to look at the phylogenetic relatedness of the species. There is a considerable amount of literature of crosses between close related species @brown2001 @arceo2016can @tong2016 but few works focused on heterospecific pollen of far related species which is a more realistic feature in natural systems due to a less likely niche overlap (REF). Although close related species seems to affect greatly species fitness @arceo2016 the effect of heterospecific pollen of far and close related species remains to be tested explicitly.

@@

The great environmental variability in natural systems and complexity of floral structures make heterospecific pollination studies a daunting task. For this reason, controlled pollination experiments where hand pollination is applied instead of pollinators as pollen vectors have simplified this task. However, other issues may arise such as overestimation of the effect (REFs), standarized methods across studies or lack of descriptive methodology for reproducible work (REFs). One of the most common ways to estimate heterospecific pollen effect through hand pollination treatments is by using 50% pollen mixes of conspecific and heterospecific pollen which a realistic proportion (REF) but challenging because of the different pollen properties such as size or hydratation levels (REFS).

>>>>

Recent studies have advanced in the ecological understanding of heterospecific pollen effect @morales2008 @ashman2013 @arceo2016. A general overview of foreign pollen arrival is that it can play an important role on species fitness but seems to be context dependent and not always produce a decrease in fitness @morales2008. Part of this unpredictability is due to the enormous variability of foreing pollen transferred in nature, where levels between 0 and 75 percent are seen, but most commonly values ranges between 0 and 20 percent of the total pollen load, being the generalist species the ones that receive greater loads of heterospecific pollen @bartomeus2008 @montgomery2012 @ashman2013 @fang2013. Although heterospecific pollen quantity is fundamental to understand the outcome of the interaction so is the different traits of both pollen donor and recipient. @ashman2013 postulated the first predictive framework for traits of heterospecific pollen effect, where different traits such as compatibility system and pollen size among others seems to be crucial to understand foreing pollen effect. Moreover, in @tong2016 an assymetric effect was shown in a crossing experiment between 6 species of the genus *Pedicularis* where the pollen of long styled species was able to grow the full length of the style on short styled species but not viceversa. Despite these recent caveats, we still lack empirical evidence to affirm what are the main traits that drive heterospecific pollen effect for both pollen donor and recipient at seed production level. Interestingly, to comprehend how these traits interact is also crucial to look at the phylogenetic relatedness of the species. There is a considerable amount of literature of crosses between close related species @brown2001 @arceo2016can @tong2016 but few works focused on heterospecific pollen of far related species which is a more realistic feature in natural systems due to a less likely niche overlap (REF). Although close related species seems to affect greatly species fitness @arceo2016 the effect of heterospecific pollen of far and close related species remains to be explored.

@@

The great environmental variability in natural systems and complexity of floral structures make heterospecific pollination studies a daunting task. For this reason, controlled pollination experiments where hand pollination is applied instead of using pollinators as pollen vectors have simplified this task. However, other issues may arise such as overestimation of the effect (REFs), different methods across studies or lack of descriptive methodology for reproducible work (REFs). One of the most common ways to estimate heterospecific pollen effect through hand pollination treatments is by using 50% pollen mixes of conspecific and heterospecific pollen @arceo2016 (More Refs).
